Studies of quantum dots

### 2016-09-14 Manuscript/paper.tex
Update paper

\newcommand{\D}{\operatorname{d\!}}
\newcommand{\ket}[1]{\left| #1 \right>}
\newcommand{\bra}[1]{\left< #1 \right|}
\newcommand{\braket}[2]{\left\langle #1 | #2\right\rangle}
\newcommand{\normord}[1]{\mathopen: #1 \mathclose:}

@@

We model the quantum dot as a collection of $N$ electrons trapped in an
external harmonic-oscillator potential.\footnote{In this paper, we use atomic
  units where $\hbar$, $m$, $e$, and $4 \pi \epsilon$ are set to 1.  Here,
  $\epsilon$ denotes the permittivity and $m$ denotes the mass, which could be
  an effective mass rather than the true electron mass.  In effect, the unit
  of energy in this article is
  $E_{\text h} (m / m_{\text e}) / (\epsilon / \epsilon_0)^2$ where
  $E_{\text h} \equiv m_{\text e} (\alpha c)^2$, the hartree unit of energy.}
Each electron possesses kinetic energy $t(\bm p) \equiv \bm p^2 / 2$ and
experiences an external potential $u(\bm r) \equiv \omega^2 \bm r^2 / 2$,
where $\omega$ is the angular frequency of the oscillator, $\bm r$ is the
2-dimensional position vector of an electron, and $\bm p$ is its linear
momentum.  Each pair of electrons experience a Coulomb repulsion
$v(\bm r, \bm r') \equiv |\bm r - \bm r'|^{-1}$ where $\bm r$ and $\bm r'$ are
their positions.  The full many-body problem is described by the
nonrelativistic Hamiltonian
\begin{align*}
  \hat H \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr) + \sum_{p < q} v(\hat{\bm r}_p, \hat{\bm r}_q)
\end{align*}
where $\hat{\bm p}_p$ and $\hat{\bm r}_p$ respectively denote the momentum and
position operators of a particle $p$.  The first summation is performed over
every particle, while the the second summation is performed over every
unordered pair of distinct particles (the ordering of $p$ is arbitrary).

The full problem is not a trivial one.  We shall first consider a much simpler
problem: a noninteracting Hamiltonian $\hat H^{\text o}$ in which the Coulomb
interaction $v$ is removed:
\begin{align} \label{eq:onebodyhamiltonian}
  \hat H^{\text o} \equiv \sum_p \hat h^{\text o}_p
  \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr)
\end{align}
In this case, the many-body Hamiltonian reduces to $N$ independent
single-particle Hamiltonians, each of the form:
\begin{align*}
  \hat h^{\text o} \equiv
  \frac{\hat{\bm p}^2 + \omega^2 \hat{\bm x}^2}{2}
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle,
whose solutions are well known.  There are various ways to construct the set
of eigenfunctions: we shall use the Fock-Darwin states that conserve angular
momentum $\hat{L}_{\text{z}}$ about the $z$-axis (perpendicular to the 2D
plane).  The eigenfunctions have the following analytic form in polar
coordinates:\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(r, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FD.R}}_{n, |m_\ell|}(\sqrt \omega r)
  \phi^{\text{FD.A}}_{m_\ell}(\varphi)
\end{gather*}
where
\begin{align*}
  \phi^{\text{FD.R}}_{n, \mu}(\rho) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + \mu)!}}
  {\mathrm e}^{- \rho^2 / 2} \rho^\mu L_n^{(\mu)}(\rho^2) \\
  \phi^{\text{FD.A}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

@@

Since Slater determinants are rather cumbersome to write out in full, we shall
instead work in the more convenient formalism of \textit{second quantization}:
\begin{align*}
  \ket{\varnothing_{p_1, \ldots, p_N}} =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger \ket{\varnothing}
\end{align*}
Here, $\varnothing$ is the vacuum state that is devoid of particles and
$\hat a_p^\dagger$ is a \textit{creation operator} for the single-particle
state $p$.  If the state $p$ is not yet occupied by a particle,
$\hat a_p^\dagger$ creates such a particle.  Otherwise, it collapses the
entire many-body state to $0$.  The Hermitian conjugate of $\hat a_p^\dagger$
is $\hat a_p$, which is an \textit{annihilation operator} for the state $p$
that has the opposite effect.  For fermions, the field operators $\hat a_p$
and $\hat a_p^\dagger$ satisfy the essential anticommutation relations,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}.
\end{align*}
Operators can be rewritten in this formalism as well.  The one-body
Hamiltonian $H^{\text o}$ defined by \eqref{eq:onebodyhamiltonian} can be
rewritten in second quantization as
\begin{align} \label{eq:second_quantized_one_body_hamiltonian}
  \hat H^{\text o} = \sum_{p, q} H_{p, q}^{\text o} \hat a_p^\dagger \hat a_q^{}
\end{align}
where the \textit{matrix element}
$H_{p, q}^{\text o} \equiv \bra{\varnothing_p} \hat H^{\text o} \ket{\varnothing_q}$.  This
can likewise be extended to $\hat V$,
\begin{align} \label{eq:second_quantized_two_body_hamiltonian}
  \hat V = \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s}^{}
  \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}
\end{align}
where $U_{p, q} \equiv \langle \varnothing_p | \hat U | \varnothing_q \rangle$ and
$V_{p, q, r, s} \equiv \langle \varnothing_{p, q} | \hat V | \varnothing_{r, s} \rangle$.
Note that due to the antisymmetry present in the Slater determinants, we have
the property that exchanging either $p$ with $q$ or $r$ with $s$ causes
$V_{p, q, r, s}$ to reverse in sign, leading to a $4$-fold symmetry that is
canceled out by the factor of $1/4$.  For this reason, the two-body matrix
element $V_{p, q, r, s}$ is commonly referred to as the \emph{antisymmetrized}
matrix element.  This is not to be confused with the non-antisymmetric
interaction integral defined by:
\begin{align*}
  \tilde V_{p, q, r, s} \equiv \int \phi_p(\bm r) \phi_q(\bm r') v(\bm r, \bm r') \phi_r(\bm r) \phi_s(\bm r') \D \bm r \D \bm r'
\end{align*}
The antisymmetrized matrix element is computed from the interaction integral
by $V_{p, q, r, s} = \tilde V_{p, q, r, s} - \tilde V_{p, q, s, r}$.  In our
work, we compute the interaction numerically via the OpenFCI software, which
numerical integrates the Coulomb interaction in the center-of-mass frame
[[cite Kvaal's paper; also, is the description correct?]].

@@

The first correction that we will apply to our solution is that of the
Hartree-Fock (HF) method.  Using the variational principle, we compute an
approximate ground state $\Psi^{\text{HF}}$ by minimizing the energy
\begin{align*}
  E^{\text{HF}} \equiv \langle \Psi^{\text{HF}} | \hat H | \Psi^{\text{HF}} \rangle
\end{align*}
with respect to $\Psi^{\text{HF}}$, subject to the restriction that
$\Psi^{\text{HF}}$ is a single Slater determinant constructed from an unknown
set of $N$ orthonormal single-particle states.  We denote each unknown
single-particle state $\psi_u$ by some label $u$.

@@

Assuming the interaction is not too strong, we can use the single-particle
states $\{\phi_p\}$ of our noninteracting Hamiltonian as a basis for the
unknown states $\{\psi_u\}$.  The two sets of states are then related by a
linear transformation matrix $\bm C$:
\begin{align*}
  \psi_u \equiv \sum_p \phi_p C_{p, u}
\end{align*}
To ensure the orthonormality of the states, we assume the set of unknown
states $\{\psi_u\}$ has as many elements as the set of $\{\phi_pp\}$ and
require the coefficient matrix $\bm C$ to be unitary.  While these conditions
are more strict than necessary, they greatly simplify the calculations and
allow the states $\{\psi_u\}$ to act as inputs for methods beyond HF.  The set
of states $\psi$ now consists of $N$ \textit{occupied} states that participate
in the Slater determinant $\Psi^{\text{HF}}$ and the remaining
\textit{unoccupied} states that do not.

Since the coefficient matrix uniquely determines the unknown states, the
variational problem has been reduced to that of solving of the coefficients
$\bm C$ that minimize the energy $E$, which is given by
\begin{align*}
  E^{\text{HF}} = \sum_{\underline u} C_{p, \underline u}^* H_{p, q}^{\text o} C_{q, \underline u}^{}
  + \sum_{\underline u, \underline v}
  C_{p, \underline u}^* C_{q, \underline v}^* V_{p, q, r, s}^{} C_{r, \underline u}^{} C_{s, \underline v}^{}
\end{align*}
where the underlined indices indicate that the summation is over occupied
states only.

Using the method of Lagrange multiplers, the minimization problem can be
reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock
  equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{\text o} + \sum_{\underline u r s}
  C_{r, \underline u}^* V_{p, r, q, s}^{} C_{s, \underline u}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element
$\epsilon_u$ associated with a particular state $u$.  [[move this sentence
later to the intro to normal-ordering: While the Fock matrix $\bm F$ may
appear to be an artificial Hamiltonian derived purely from the minimization
problem, it can be considered the normal-ordered operator with respect to a
non-vacuum state, as explained later.]]

@@

The second-quantized operators in
\eqref{eq:second_quantized_one_body_hamiltonian} and
\eqref{eq:second_quantized_two_body_hamiltonian} are constructed with respect
to the vacuum state.  That is, the operator strings
$\hat{a}_p^\dagger \hat{a}_q$ and
$\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have a vacuum
expectation value of zero:
\begin{align*}
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q \ket{\varnothing} = 0 &
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{\varnothing} = 0
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state
$\varnothing$.  It is possible to be normal-ordered with respect to a
non-vacuum state, however.  Consider a Slater determinant $\Phi$ with
$N_{\text F}$ occupied states.  One can define a set of \textit{quasiparticle
  field operators} $\hat b$ and $\hat b^\dagger$ that obey the same
anticommutation relations as the $\hat a$ operators.  The field operators are
defined as:
\begin{align*}
  \hat b_p \equiv \hat a_{\overline p} + \hat a_{\underline p}^\dagger = \begin{cases}
    \hat a_{\overline p} & \text{if $\overline p$ is unoccupied in $\Phi$} \\
    \hat a_{\underline p}^\dagger & \text{if $\underline p$ is occupied in $\Phi$}
  \end{cases}
\end{align*}
Here we use an underline to indicate that only an occupied state is accepted
and an overline to indicate only an unoccupied state is accepted.  If the
state is not acceptable, then that term does not contribute to the overall
expression.

We can then define normal ordering with respect to the reference state $\Phi$:
namely, rearrange an operator string such that every creation operator appears
before an annihilation operator by repeatedly exchanging neighboring
operators, with each exchange costing a minus sign.  In this form, it is
guaranteed that the expectation value of the operator string with respect to
$\Phi$ is zero:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
We denote normal ordering of an operator string with respect to $\Phi$ by a
pair of colons:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
It is not particularly interesting to examine normal ordering of the
quasiparticle field operators, since the result is trivial to compute.  What
is more interesting is the normal ordering of the physical field operators
$\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} =
  -\hat{a}_{\underline q} \hat{a}_{\underline p}^\dagger
  + \hat{a}_{\underline p}^\dagger \hat{a}_{\overline q}
  + \hat{a}_{\overline p}^\dagger \hat{a}_{\underline q}
  + \hat{a}_{\overline p}^\dagger \hat{a}_{\overline q} =
  -\delta_{\underline q \underline p}
  + \hat{a}_p^\dagger \hat{a}_q
\end{align*}
This makes the normal ordering notation a useful and concise way of expressing
the right-hand side.

Using the normal ordering formalism, we can rewrite the Hamiltonian
$\hat H = \hat H^{\text o} + \hat V$ as
\begin{align*}
  \hat{H} = E + \sum_{p, q} F_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
\end{align*}
where
\begin{align*}
  E &= \sum_{\underline p} H^{\text o}_{\underline p, \underline p} + \frac{1}{2} \sum_{\underline p, \underline q} V_{\underline p, \underline q, \underline p, \underline q} \\
  F_{p, q} &= H^{\text o}_{\underline p, \underline q} + \sum_{\underline r} V_{\underline p, \underline r, \underline q, \underline r} \\
\end{align*}
Notice that $E$ is in fact the Hartree-Fock energy and $F_{p, q}$ is the Fock matrix.

@@

The commutator in the flow equations (\ref{eq:imsrgode}) guarantees that the
IM-SRG wave function $U^\dagger(s) \ket{\Phi}$ can be expanded in terms of
linked diagrams only \cite{shavitt2009many,ISI:A1981MN73700014}, which
suggests that IM-SRG is size-extensive. Regarding the quality of the SRG
results, it means that the error introduced by truncating the many-body
expansions scales linearly with the number of particles $N$.

>>>>

\newcommand{\D}{\operatorname{d\!}}
\newcommand{\ket}[1]{| #1 \rangle}
\newcommand{\bra}[1]{\langle #1 |}
\newcommand{\braket}[2]{\langle #1 | #2\rangle}
\newcommand{\normord}[1]{\mathopen: #1 \mathclose:}

@@

We model the quantum dot as a collection of $N$ electrons trapped in an
external harmonic-oscillator potential.\footnote{In this paper, we use atomic
  units where $\hbar$, $m$, $e$, and $4 \pi \epsilon$ are set to 1.  Here,
  $\epsilon$ denotes the permittivity and $m$ denotes the mass, which could be
  an effective mass rather than the true electron mass.  In effect, the unit
  of energy in this article is
  $E_{\text h} (m / m_{\text e}) / (\epsilon / \epsilon_0)^2$ where
  $E_{\text h} \equiv m_{\text e} (\alpha c)^2$, the hartree unit of energy.}
Each electron possesses kinetic energy $t(\bm p) \equiv \bm p^2 / 2$ and
experiences an external potential $u(\bm r) \equiv \omega^2 \bm r^2 / 2$,
where $\omega$ is the angular frequency of the oscillator, $\bm r$ is the
2-dimensional position vector of an electron, and $\bm p$ is its linear
momentum.  Each pair of electrons experience a Coulomb repulsion
$v(\bm r, \bm r') \equiv |\bm r - \bm r'|^{-1}$ where $\bm r$ and $\bm r'$ are
their positions.  The full many-body problem is described by the
nonrelativistic Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{\text o} + \hat V = \sum_{p} \bigl(
    t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr) + \sum_{p < q} v(\hat{\bm r}_p, \hat{\bm r}_q)
\end{align*}
where $\hat{\bm p}_p$ and $\hat{\bm r}_p$ respectively denote the momentum and
position operators of a particle $p$.  The first summation is performed over
every particle, while the the second summation is performed over every
unordered pair of distinct particles (the ordering of $p$ is arbitrary).

The full problem is not a trivial one.  We shall first consider a much simpler problem: the noninteracting part of the Hamiltonian, $\hat H^{\text o}$, which does not contain the Coulomb interaction $v$:
\begin{align} \label{eq:onebodyhamiltonian}
  \hat H^{\text o} \equiv \sum_p \hat h^{\text o}_p
  \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr)
\end{align}
In this case, the many-body Hamiltonian reduces to $N$ independent
single-particle Hamiltonians, each of the form:
\begin{align*}
  \hat h^{\text o} \equiv
  \frac{\hat{\bm p}^2 + \omega^2 \hat{\bm x}^2}{2}
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle,
whose solutions are well known.  There are various ways to construct the set
of eigenfunctions: we shall use the Fock-Darwin states that conserve angular
momentum $\hat{L}_{\text{z}}$ about the $z$-axis (perpendicular to the 2D
plane).  The eigenfunctions have the following analytic form in polar
coordinates:\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(r, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FD.R}}_{n, |m_\ell|}(\sqrt \omega r)
  \phi^{\text{FD.A}}_{m_\ell}(\varphi)
\end{gather*}
where
\begin{align*}
  \phi^{\text{FD.R}}_{n, \mu}(\rho) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + \mu)!}}
  {\mathrm e}^{- \rho^2 / 2} \rho^\mu L_n^{(\mu)}(\rho^2) \\
  \phi^{\text{FD.A}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

@@

Since Slater determinants are rather cumbersome to write out in full, we shall
instead work in the more convenient formalism of \textit{second quantization}:
\begin{align*}
  \ket{\varnothing_{p_1, \ldots, p_N}} =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger \ket{\varnothing}
\end{align*}
Here, $\varnothing$ is the vacuum state that is devoid of particles and
$\hat a_p^\dagger$ is a \textit{creation operator} for the single-particle
state $p$.  If the state $p$ is not yet occupied by a particle,
$\hat a_p^\dagger$ creates such a particle.  Otherwise, it collapses the
entire many-body state to $0$.  The Hermitian conjugate of $\hat a_p^\dagger$
is $\hat a_p$, which is an \textit{annihilation operator} for the state $p$
that has the opposite effect.  For fermions, the field operators $\hat a_p$
and $\hat a_p^\dagger$ satisfy the essential anticommutation relations,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = [p = q].
\end{align*}
where $[P]$ denotes the Iverson bracket, which returns $1$ if the predicate $P$ is true and $0$ otherwise.  Here, $[p = q]$ is simply an alternative way to denote the Kronecker delta $\delta_{p, q}$, but this notation will be used later for other kinds of predicates as well.

Operators can be rewritten in this formalism as well.  The one-body
Hamiltonian $H^{\text o}$ defined by \eqref{eq:onebodyhamiltonian} can be
rewritten in second quantization as
\begin{align} \label{eq:second_quantized_one_body_hamiltonian}
  \hat H^{\text o} = \sum_{p, q} H_{p, q}^{\text o} \hat a_p^\dagger \hat a_q^{}
\end{align}
where the \textit{matrix element}
$H_{p, q}^{\text o} \equiv \bra{\varnothing_p} \hat H^{\text o} \ket{\varnothing_q}$.  This
can likewise be extended to $\hat V$,
\begin{align} \label{eq:second_quantized_two_body_hamiltonian}
  \hat V = \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s}^{}
  \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}
\end{align}
where $U_{p, q} \equiv \langle \varnothing_p | \hat U | \varnothing_q \rangle$ and
$V_{p, q, r, s} \equiv \langle \varnothing_{p, q} | \hat V | \varnothing_{r, s} \rangle$.
The complete Hamiltonian is given by their sum:
\begin{align}
  \hat{H} = \sum_{p, q} H^{\text o}_{p, q} \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s  \label{eq:vacuumhamiltonian}
\end{align}
Note that due to the antisymmetry present in the Slater determinants, we have
the property that exchanging either $p$ with $q$ or $r$ with $s$ causes
$V_{p, q, r, s}$ to reverse in sign, leading to a $4$-fold symmetry that is
canceled out by the factor of $1/4$.  For this reason, the two-body matrix
element $V_{p, q, r, s}$ is commonly referred to as the \emph{antisymmetrized}
matrix element.  This is not to be confused with the non-antisymmetric
interaction integral defined by:
\begin{align*}
  \tilde V_{p, q, r, s} \equiv \int \phi_p(\bm r) \phi_q(\bm r') v(\bm r, \bm r') \phi_r(\bm r) \phi_s(\bm r') \D \bm r \D \bm r'
\end{align*}
The antisymmetrized matrix element is computed from the interaction integral
by $V_{p, q, r, s} = \tilde V_{p, q, r, s} - \tilde V_{p, q, s, r}$.  In our
work, we compute the interaction numerically via the OpenFCI software, which
numerical integrates the Coulomb interaction in the center-of-mass frame
[[cite Kvaal's paper; also, is the description correct?]].

@@

The first correction that we will apply to our solution is that of the
Hartree-Fock (HF) method.  Using the variational principle, we compute an
approximate ground state $\Psi_{\text{HF}}$ by minimizing the energy
\begin{align*}
  E_{\Psi_{\text{HF}}} \equiv \langle \Psi_{\text{HF}} | \hat H | \Psi_{\text{HF}} \rangle
\end{align*}
with respect to $\Psi_{\text{HF}}$, subject to the restriction that
$\Psi_{\text{HF}}$ is a single Slater determinant constructed from an unknown
set of $N$ orthonormal single-particle states.  We denote each unknown
single-particle state $\psi_u$ by some label $u$.

@@

Assuming the interaction is not too strong, we can use the single-particle
states $\{\phi_p\}$ of our noninteracting Hamiltonian as a basis for the
unknown states $\{\psi_u\}$.  The two sets of states are then related by a
linear transformation matrix $\bm C$:
\begin{align*}
  \psi_u \equiv \sum_p \phi_p C_{p, u}
\end{align*}
To ensure the orthonormality of the states, we assume the set of unknown
states $\{\psi_u\}$ has as many elements as the set of $\{\phi_p\}$ and
require the coefficient matrix $\bm C$ to be unitary.  While these conditions
are more strict than necessary, they greatly simplify the calculations and
allow the states $\{\psi_u\}$ to act as inputs for methods beyond HF.  The set
of states $\psi$ now consists of $N$ \textit{occupied} states that participate
in the Slater determinant $\Psi_{\text{HF}}$ and the remaining
\textit{unoccupied} states that do not.

Since the coefficient matrix uniquely determines the unknown states, the
variational problem has been reduced to that of solving of the coefficients
$\bm C$ that minimize the energy $E_{\Psi_{\text{HF}}}$, which is given by
\begin{align}
  E_{\Psi_{\text{HF}}} &= \sum_{u \in \text{occ.}} C_{p, u}^* H_{p, q}^{\text o} C_{q, u}^{}
  \notag \\ &\hphantom{{}=} + \sum_{u, v \in \text{occ.}}
  C_{p, u}^* C_{q, v}^* V_{p, q, r, s}^{} C_{r, u}^{} C_{s, v}^{} \label{eq:hfenergy}
\end{align}

Using the method of Lagrange multiplers, the minimization problem can be
reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock
  equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{\text o} + \sum_{r, s \in \text{any}; u \in \text{occ.}}
  C_{r, u}^* V_{p, r, q, s}^{} C_{s, u}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element
$\epsilon_u$ associated with a particular state $\psi_u$.  [[move this sentence
later to the intro to normal-ordering: While the Fock matrix $\bm F$ may
appear to be an artificial Hamiltonian derived purely from the minimization
problem, it can be considered the normal-ordered operator with respect to a
non-vacuum state, as explained later.]]

@@

The second-quantized operators in
\eqref{eq:second_quantized_one_body_hamiltonian} and
\eqref{eq:second_quantized_two_body_hamiltonian} are constructed with respect
to the vacuum state.  That is, the operator strings
$\hat{a}_p^\dagger \hat{a}_q$ and
$\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have a vacuum
expectation value of zero:
\begin{align*}
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q \ket{\varnothing} = 0 &
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{\varnothing} = 0
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state
$\varnothing$.  It is possible to be normal-ordered with respect to a
non-vacuum state, however.  Consider a Slater determinant $\Phi$ with
$N_{\text F}$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{\varnothing_{i_1, i_2, \ldots, i_{N_{\text F}}}}
\end{align*}
Here $\{i_1, i_2, \ldots, i_{N_{\text F}}\}$ is a set of $N_{\text F}$ indices that label the occupied states.

One can define a set of \textit{quasiparticle
  field operators} $\hat b$ and $\hat b^\dagger$ that obey the same
anticommutation relations as the $\hat a$ operators.  The field operators are
defined as:
\begin{align*}
  \hat b_p \equiv \begin{cases}
    \hat a_{p} & \text{if $p$ is unoccupied in $\Phi$} \\
    \hat a_{p}^\dagger & \text{if $p$ is occupied in $\Phi$}
  \end{cases}
\end{align*}
The quasiparticle field operators consider $\ket{\Phi}$ to be their ``vacuum state'' (sometimes called ``Fermi vaccuum''), even though it is not a true vacuum state.  Analogously, it is possible to construct any other Slater determinant by applying quasiparticle operators on top of the reference state:
\begin{align*}
  \ket{\Phi_{p_1, \ldots, p_k}} \equiv \hat{b}^\dagger_{p_1} \cdots \hat{b}^\dagger_{p_k} \ket{\Phi}
\end{align*}
In nonrelativistic many-body theory, it is expected that $\{p_1, \ldots, p_k\}$ contain a balanced number of occupied and unoccupied states to conserve the total number of particles.

We can then define normal ordering with respect to the reference state $\Phi$: namely, rearrange an operator string such that every quasiparticle creation operator appears before a quasiparticle annihilation operator by repeatedly exchanging neighboring operators, with each exchange costing a minus sign.  In this form, it is guaranteed that the expectation value of the operator string with respect to
$\Phi$ is zero:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
We denote normal ordering of an operator string with respect to $\Phi$ by a
pair of colons:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
It is not particularly interesting to examine normal ordering of the
quasiparticle field operators, since the result is trivial to compute.  What
is more interesting is the normal ordering of the physical field operators
$\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} &=
  \hat{a}_p^\dagger \hat{a}_q - [p, q \in \text{occ.} \text{ and } p = q]
\end{align*}
This makes the normal ordering notation a useful and concise way of expressing the right-hand side.

Using the normal ordering formalism, we can rewrite the Hamiltonian
$\hat H = \hat H^{\text o} + \hat V$ as
\begin{align}
  \hat{H} = E_\Phi + \sum_{p, q} F_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  E_\Phi &= \sum_{i \in \text{occ.}} H^{\text o}_{i, i} + \frac{1}{2} \sum_{i, j \in \text{occ.}} V_{i, j, i, j} \\
  F_{p, q} &= H^{\text o}_{p, q} + \sum_{i \in \text{occ.}} V_{p, i, q, i} \\
\end{align*}
Notice that $E_\Phi$ is in fact the Hartree-Fock energy and $F_{p, q}$ is the Fock matrix.  The representation of $\hat H$ in \eqref{eq:normordhamiltonian} is completely equivalent to that of $\hat H$ in \eqref{eq:vacuumhamiltonian}.  The only difference is that it shuffles the matrix elements so as to put the reference state $\ket{\Phi}$ into the main focus, as opposed to the vacuum state $\ket{\varnothing}$.  Where this makes a crucial difference is when the equations become \emph{truncated}.

@@

The commutator in the flow equations \eqref{eq:imsrgode} guarantees that the
IM-SRG wave function $\hat U(s) \ket{\Phi}$ can be expanded in terms of
linked diagrams only \cite{shavitt2009many,ISI:A1981MN73700014}, which
suggests that IM-SRG is size-extensive.  Regarding the quality of the SRG
results, it means that the error introduced by truncating the many-body
expansions scales linearly with the number of particles $N$.

With this truncation, the generator $\hat{\eta}$ can be written as
\begin{align*}
\hat{\eta} = \sum_{p, q} \eta_{p, q}^{[1]} \normord{\hat a_p^\dagger \hat a_q} +
\frac{1}{4} \sum_{p, q, r, s}\eta_{p, q, r, s}^{[2]} \normord{\hat a_p^\dagger \hat a_q^\dagger \hat a_s \hat a_r}
\end{align*}
where $\eta_{p, q}^{[1]}$ and $ \eta_{p, q, r, s}^{[2]}$ are the one- and two-body elements respectively.  For consistency, we define $H^{[0]} \equiv E_\Phi$, $H^{[1]}_{p, q} \equiv F_{p, q}$, and $H^{[2]}_{p, q, r, s} \equiv V_{p, q, r, s}$ as well.
